Nonlinear methods to quantify Movement Variability in Human-Humanoid Interaction Activities

### 2021-03-18 docs/manuscript/main.tex
updates references for github to srep2021
fixes references for figure01

Fig~\ref{fig:ssr} illustrates the state space reconstruction.

@@

	Code and data to reproduce the figure is available in \cite{srep2020}.

@@

    \label{fig:ssr}

@@

As an example of the use of $E_1(m)$ and $E_2(m)$ values, we consider two time 
series: the solution for the $x$ variable of the Lorenz system 
(Fig~\ref{fig:caoami}A), and a Gaussian noise time series with zero mean 
and a variance of one (Fig~\ref{fig:caoami}B).
We then compute $E_1(m)$ and $E_2(m)$ values for each time series.
The $E_1(m)$ values for the chaotic time series appear to be constant
after the dimension is equal to six.
The determination of six is given that any value of $m$ can be used as they 
are within the threshold of $1\pm0.05$ (Fig~\ref{fig:caoami}C).
$E_2(m)$ values, for chaotic time series, are different to one 
(Fig~\ref{fig:caoami}E), for which, it can be concluded that for the 
chaotic time series the minimum embedding dimension the time series 
comes from a deterministic signal. With regard to the noise time series,  
$E_1(m)$ values appeared to be constant when $m$ is close to thirteen, 
which is defined by the threshold of $1\pm0.05$ (Fig~\ref{fig:caoami}D).
$E_1(m)$ values then indicate the minimum embedding dimension of the 
noisy time series is thirteen, however all of the $E_2(m)$ values are 
approximately equal to one (Figure~\ref{fig:caoami}F) for which it can be 
concluded that noise time series is a stochastic signal.
It is important to note that for this work not only $E_1(m)$ and $E_2(m)$ are 
computed but also a variation of $\tau$ from 1 to 20 is presented. 
The purpose of such variation for $\tau$ is to show its independence with
regard to $E_1(m)$ and $E_2(m)$ values as $\tau$ is increasing 
(Fig~\ref{fig:caoami}C,D,E and F). 
However, one negative of the Cao's algorithm \cite{Cao1997} is the definition of 
a new threshold where $m$ values appear to be constant in $E_1 (m)$.
In the case of the given examples and reported results, we defined such 
threshold as 0.05. Further investigation is required for the selection of the 
threshold in the $E_1(m)$, as the selection of the threshold in this work is 
base on no particular method but visual inspection.

@@

For example, we compute the AMI for two time series:
A) the $x$ solution of the Lorenz system, and
B) a noise time series using a normal distribution with mean zero and 
standard deviation equal to one. From Fig~\ref{fig:caoami}(G, H), it can then be 
concluded that the amount of knowledge for any noise time series is zero 
for which the first minimum embedding parameter is $\tau_0=1$. 
On the contrary, the first minimum of the AMI for the chaotic time series 
is $\tau_0=17$ which is the value that maximize the independence 
between $x(n)$ and $x(n+\tau)$ in the reconstructed state space 
\cite{bradley2015}.
%%---------------------------------(FIGURE)-------------------------------------
Similarly as Cao's algorithm negatives, AMI's algorithm is not an
exception for negatives, which are worthwhile to mention for further 
investigations.
For instance, (i) is not clear why the choose of the first minimum of the AMI 
is the minimum delay embedding parameter \cite{kantz2003} and 
(ii) the probability distribution of the AMI function is computed
with the use of histograms which depends on a heuristic choice of number of bins
for which AMI depends on partitioning \cite{garcia2005e71}.

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/methods/caoami/versions/drawing-v00.pdf}
%    \caption{
%	{\bf Minimum delay embedding values with AMI's method and minimum dimension embedding values with Cao's method.} 
%	(A) chaotic and (B) random time series.
%	(C, D) $E_1 (m)$ values and (E, F) $E_2(m)$ values 
%	with variations of $\tau$ values from one to twenty
%	for chaotic and random time series.
%	(G, H) AMI values where its first minimum value in the curve
%	is the minimum time delay embedding ($\tau_0$).
%	Code and data to reproduce the figure is available in \cite{srep2020}.
%        }
%    \label{fig:caoami}
%\end{figure}
%%---------------------------------(FIGURE)-------------------------------------

@@

For the creation of a recurrence plot based on time series 
$\{ \boldsymbol{x}_n \}$, it is first computed the state space 
reconstruction with uniform time-delay embedding 
$X(i)=\{ \boldsymbol{ \tilde{x} }_n, \dots,  
\boldsymbol{ \tilde{x} }_{n -(m-1)\tau} \}$
where $i=1,\dots,N$, $N$ is the number of considered states of $X(i)$ 
and $X(i) \in \mathbb{R}^m$ \cite{eckmann1987}.
The recurrence plot is therefore a two-dimensional $N \times N$ square 
matrix, $\mathbf{R}$, where a black dot is placed at $(i,j)$ 
whenever $X(i)$ is sufficiently close to $X(j)$: 
%%********************************[EQUATION]************************************
\begin{equation}
\mathbf{R}^{m}_{i,j} (\epsilon) = \Theta ( \epsilon_i - || X(i) - X(j) ||
\end{equation}
%%********************************[EQUATION]************************************
where $\quad i,j=1,\dots,N$, $\epsilon$ is a threshold distance, 
$|| \cdotp ||$ a norm, and $\Theta(\cdotp)$ is the Heaviside 
function (i.e. $\Theta(x)=0$, if $x<0$, and $\Theta(x)=1$ otherwise) 
(Fig~\ref{fig:rps}(A,B)) \cite{eckmann1987, marwan2007,marwan2015}.
RP is also characterised with a line of identity (LOI) which is a  
diagonal line due to $ R_{i,j}=1 (i,j=1,\dots,N)$.

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\subsection*{Structures of Recurrence Plots}
Pattern formations in the RPs can be designated either 
as topology for large-scale patterns or texture for small-scale patterns.
In the case of topology, the following pattern formations are commonly presented:
(i) homogeneous where uniform recurrence points are spread in the RP e.g., 
uniformly distributed noise (Fig~\ref{fig:rps}C), 
(ii) periodic and quasi-periodic systems where diagonal lines and 
checkerboard structures represent oscillating systems, e.g., sinusoidal 
signals (Fig~\ref{fig:rps}D), 
(iii) drift where paling or darkening recurrence points away from 
the LOI is caused by drifting systems, 
e.g., logistic map (Fig~\ref{fig:rps}E), and
(iv) disrupted where recurrence points are presented white areas or 
bands that indicate abrupt changes in the dynamics, e.g. Brownian motion 
(Fig~\ref{fig:rps}F) \cite{eckmann1987, marwan2015}.
Texture patterns in RPs can be categorised as:
(i) single or isolated recurrence points that represent rare occurring states, 
and do not persist for any time or fluctuate heavily,
(ii) dots forming diagonal lines where the length of the small-scale parallel 
lines in the diagonal are related to the ratio of determinism or predictability 
in the dynamics of the system, and
(iii) dots forming vertical and horizontal lines where the length of the 
lines represent a time length where a state does not change or change very 
slowly and these patterns formation represent discontinuities in the signal, 
and (iv) dots clustering to inscribe rectangular regions which are related 
to laminar states or singularities \cite{marwan2015}.

@@

%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/methods/rps/versions/drawing-v00.pdf}
%    \caption{
%	{\bf Recurrence Plots and its patterns.} 
%	(A) State space of the Lorenz system with controlling parameters 
%	($\rho=28, \sigma=10, \beta=8/3$) where the point, $j$, in trajectory $X()$ which falls into the neighborhood 
%	(black circle) of a given point at $i$ is a recurrent point and is 
%	represented as a black dot in the recurrence plot at location 
%	$(i, j)$ or white otherwise.
%	(B) Recurrence plot using the 
%	three components of the Lorenz system and the RP with no embeddings 
%	and threshold $\epsilon=5$.
%	Time-series with its respective recurrence plot for:
%	(C) uniformly distributed noise,
%	(D) super-positionet harmonic oscillation 
%	($sin( \frac{1}{5}*t) * sin( \frac{5}{100}*t) $),
%	(E) drift logistic map ($x_{i+1} = 4 x_i (1- x_i) $) corrupted 
%	with a linearly increase term ($0.01*i$), and
%	(F) disrupted brownian motion  ($x_{i+1} = x_i + 2*rnorm(1) $).
%	This figure is adapted from \cite{marwan2015}.
%	Code and data to reproduce the figure is available in \cite{srep2020}.
%	}
%    \label{fig:rps}
%\end{figure}
%%%---------------------------------(FIGURE)-------------------------------------

@@

\subsection*{Human-humanoid imitation activities}
For human-humanoid imitation (HHI) activities four neMEMSi sensors were used,
two of which were attached to the right hand of the participant and the 
other two to the left hand of the humanoid robot.
Then, each participant was asked to imitate repetitions of simple horizontal
and vertical arm movements performed by the humanoid robot in the following 
conditions:
(i) ten repetitions of horizontal arm movement at normal (HN) 
	and faster (HF) speed (Figure~\ref{fig:hri} A), and
(ii) ten repetitions of vertical arm movement at normal (VN) 
	and faster (VF) speed (Figure~\ref{fig:hri} C).
The normal and faster speed of arm movements is defined by the duration 
in number of samples of one repetition of NAO's arm movements. 
We select NAO's arm movements duration to distinguish between normal and 
faster arm movements as NAO's movements have less variation 
for such repetitive movements. 
The duration for one repetition of the horizontal 
arm movement at normal speed, HN, is 5 seconds considering 
that each repetition last around 250 samples.
For horizontal arm movement at faster speed, HF, each repetition were performed 
in 2 seconds which correspond to 90 samples of data.
The vertical arm movement at normal speed, VN, were performed  in 6 seconds 
which is around 300 samples of data. For vertical arm movement at 
faster speed, VF, each repetition lasts about 2.4 seconds which correspond 
to 120 samples of data.
To visualise the distinction between normal and faster speed for horizontal 
and vertical arm movements, Fig~\ref{fig:sts} shows smoothed time series 
for axes Z and Y of the gyroscope sensors with four window lengths: 
2-sec (100-samples), 5-sec (250-samples), 10-sec (500-samples) 
and 15-sec (750-samples).
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht] 
\centering
\includegraphics[width=1.0\textwidth]{figures/experiment/sts/pdf/sts}
    \caption{
	{\bf Time series duration of horizontal and vertical arm movements.} 
    		Time series of smoothed data from gyroscope sensor 
		for different speed arm movements performed by NAO: 
		(A) Horizontal Normal arm movement (HN), 
		(B) Horizontal Faster arm movement (HF),
		(C) Vertical Normal arm movement (VN) and 
		(D) Vertical Faster arm movement (VF).
		Additionally, 
		(A) shows window sizes for 2-seconds (100 samples), 
		5-seconds (250 samples), 10-seconds (500 samples) 
		and 15-seconds (750 samples)
		which are also presented in (B), (C) and (D).
		Code and data to reproduce the figure is available in \cite{srep2020}.
        }
	\label{fig:sts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Data collection from inertial measurement units} 
\label{sec:experiment:subsec:imu}
To give insight to the research questions, 
we considered various conditions of time series collected for this work: 
\begin{itemize}
\item Three levels of smoothness for normalised data 
	(e.g., sg0zmuv, sg1zmuv and sg2zmuv) where sg 
	stands for Savitzky-Golay filter with two different filter lengths (29 and 159) 
	and the same polynomial degree of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal}
	and zmuv is zero mean unit variance.
\item four arm movement activities with two velocities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}
%After the data collection, raw time series were windowed, normalised and smoothed. 
Due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig:ts}).
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
    	\caption{ 
	{\bf Time series for horizontal and vertical arm movements.}
		(A/D) raw-normalised (sg0zmuv), 
		(B/E) normalised-smoothed 1 (sg1zmuv) and
		(C/F) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants 
		($p01$, $p02$, and $p03$) 
		for horizontal and vertical arm movements in normal 
		and faster velocity (HN, HF, VN, VF) 
		with the normalised GyroZ or GyroY axis 
		and with one sensor attached to the participant (HS01) 
		and other sensor attached to the robot (RS01).	
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:ts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

@@

%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/caoami/pdf/fig2}
	\caption{
	{\bf Box plots of minimum embedding parameters.} 
	Box plots of (A) minimum embedding dimensions 
	and (B) first minimum AMI values for 
	Horizontal Normal (HN), Horizontal Faster (HF),
	Vertical Normal (VN) and Vertical Faster (VF)
	with sensors attached to participants (HS01) and
	sensor attached to robot (RS01).
	Minimum embedding parameters are for twenty participants 
	($p01$ to $p20$) with three smoothed signals 
	(sg0: sg0zmuvGyroZ, sg1: sg1zmuvGyroZ and sg2: sg2zmuvGyroZ)
	and window length of 10-sec (500 samples).
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:cao_ami}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsubsection*{Uniform Time-Delay Embedding}
Using the overall embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$), 
the first three axis of the rotated Principal Components Analysis (PCA) 
are shown for the reconstructed state spaces of horizontal (figures~\ref{fig:rsss}(A)) 
and vertical (figures~\ref{fig:rsss}(B)) arm movements. 
While visual inspection of these figures 
suggests differences in the trajectories of the reconstructed state spaces, 
we require an objective quantification to determine the extent of the differences.  
One approach could be use Euclidean distances from the origin for points in these 
trajectories, but this proved inconclusive and was not able to capture 
the dynamics of the trajectories. Consequently, we applied 
Recurrence Plots and Recurrence Quantification Analysis.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/versions/drawing-v00.pdf}
\caption{
	{\bf RSSs for horizontal and vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$
	for (A) horizontal and (B) vertical arm movements.
	Code and data to reproduce the figure is available in \cite{srep2020}.	
        }
    \label{fig:rsss}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrences Plots}
Using the average embedding parameters 
($\overline{m}_0=6$, $\overline{\tau}_0=8$) 
and an recurrence threshold of $\epsilon=1$.
As our interest is for dynamical transitions, 
there is little importance on the selection of $\epsilon$ which in this case
is 1. 
Recurrence Plots (RP) were computed for horizontal (figures~\ref{fig:rpsr}(A)) and 
vertical (figures~\ref{fig:rpsr}(B)) arm movements.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/versions/drawing-v00.pdf}
\caption{
	{\bf RPs for horizontal (A) and vertical (B) arm movements.}	
	Recurrence plots were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$.
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:rpsr}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

>>>>

Figures~\ref{fig01}(A.1 to D.1) illustrates the state space reconstruction.

@@

	Code and data to reproduce the figure is available in \cite{srep2021}.

@@

    \label{fig01}

@@

As an example of the use of $E_1(m)$ and $E_2(m)$ values, we consider two time 
series: the solution for the $x$ variable of the Lorenz system 
(Fig~\ref{fig01}A.2), and a Gaussian noise time series with zero mean
and a variance of one (Fig~\ref{fig01}B.2).
We then compute $E_1(m)$ and $E_2(m)$ values for each time series.
The $E_1(m)$ values for the chaotic time series appear to be constant
after the dimension is equal to six.
The determination of six is given that any value of $m$ can be used as they 
are within the threshold of $1\pm0.05$ (Fig~\ref{fig01}C.2).
$E_2(m)$ values, for chaotic time series, are different to one 
(Fig~\ref{fig01}E.2), for which, it can be concluded that for the
chaotic time series the minimum embedding dimension the time series 
comes from a deterministic signal. With regard to the noise time series,  
$E_1(m)$ values appeared to be constant when $m$ is close to thirteen, 
which is defined by the threshold of $1\pm0.05$ (Fig~\ref{fig01}D.2).
$E_1(m)$ values then indicate the minimum embedding dimension of the 
noisy time series is thirteen, however all of the $E_2(m)$ values are 
approximately equal to one (Figure~\ref{fig01}F.2) for which it can be
concluded that noise time series is a stochastic signal.
It is important to note that for this work not only $E_1(m)$ and $E_2(m)$ are 
computed but also a variation of $\tau$ from 1 to 20 is presented. 
The purpose of such variation for $\tau$ is to show its independence with
regard to $E_1(m)$ and $E_2(m)$ values as $\tau$ is increasing 
(Fig~\ref{fig01}C.2,D.2,E.2 and F.2).
However, one negative of the Cao's algorithm \cite{Cao1997} is the definition of 
a new threshold where $m$ values appear to be constant in $E_1 (m)$.
In the case of the given examples and reported results, we defined such 
threshold as 0.05. Further investigation is required for the selection of the 
threshold in the $E_1(m)$, as the selection of the threshold in this work is 
base on no particular method but visual inspection.

@@

For example, we compute the AMI for two time series:
A) the $x$ solution of the Lorenz system, and
B) a noise time series using a normal distribution with mean zero and 
standard deviation equal to one. From Fig~\ref{fig01}(G.2, H.2), it can then be
concluded that the amount of knowledge for any noise time series is zero 
for which the first minimum embedding parameter is $\tau_0=1$. 
On the contrary, the first minimum of the AMI for the chaotic time series 
is $\tau_0=17$ which is the value that maximize the independence 
between $x(n)$ and $x(n+\tau)$ in the reconstructed state space 
\cite{bradley2015}.
%%---------------------------------(FIGURE)-------------------------------------
Similarly as Cao's algorithm negatives, AMI's algorithm is not an
exception for negatives, which are worthwhile to mention for further 
investigations.
For instance, (i) is not clear why the choose of the first minimum of the AMI 
is the minimum delay embedding parameter \cite{kantz2003} and 
(ii) the probability distribution of the AMI function is computed
with the use of histograms which depends on a heuristic choice of number of bins
for which AMI depends on partitioning \cite{garcia2005e71}.

%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/methods/caoami/versions/drawing-v00.pdf}
%    \caption{
%	{\bf Minimum delay embedding values with AMI's method and minimum dimension embedding values with Cao's method.} 
%	(A) chaotic and (B) random time series.
%	(C, D) $E_1 (m)$ values and (E, F) $E_2(m)$ values 
%	with variations of $\tau$ values from one to twenty
%	for chaotic and random time series.
%	(G, H) AMI values where its first minimum value in the curve
%	is the minimum time delay embedding ($\tau_0$).
%	Code and data to reproduce the figure is available in \cite{srep2021}.
%        }
%    \label{fig:caoami}
%\end{figure}
%%---------------------------------(FIGURE)-------------------------------------

@@

For the creation of a recurrence plot based on time series 
$\{ \boldsymbol{x}_n \}$, it is first computed the state space 
reconstruction with uniform time-delay embedding 
$X(i)=\{ \boldsymbol{ \tilde{x} }_n, \dots,  
\boldsymbol{ \tilde{x} }_{n -(m-1)\tau} \}$
where $i=1,\dots,N$, $N$ is the number of considered states of $X(i)$ 
and $X(i) \in \mathbb{R}^m$ \cite{eckmann1987}.
The recurrence plot is therefore a two-dimensional $N \times N$ square 
matrix, $\mathbf{R}$, where a black dot is placed at $(i,j)$ 
whenever $X(i)$ is sufficiently close to $X(j)$: 
%%********************************[EQUATION]************************************
\begin{equation}
\mathbf{R}^{m}_{i,j} (\epsilon) = \Theta ( \epsilon_i - || X(i) - X(j) ||
\end{equation}
%%********************************[EQUATION]************************************
where $\quad i,j=1,\dots,N$, $\epsilon$ is a threshold distance, 
$|| \cdotp ||$ a norm, and $\Theta(\cdotp)$ is the Heaviside 
function (i.e. $\Theta(x)=0$, if $x<0$, and $\Theta(x)=1$ otherwise) 
(Fig~\ref{fig01}(A.3,B.3)) \cite{eckmann1987, marwan2007,marwan2015}.
RP is also characterised with a line of identity (LOI) which is a  
diagonal line due to $ R_{i,j}=1 (i,j=1,\dots,N)$.

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\subsection*{Structures of Recurrence Plots}
Pattern formations in the RPs can be designated either 
as topology for large-scale patterns or texture for small-scale patterns.
In the case of topology, the following pattern formations are commonly presented:
(i) homogeneous where uniform recurrence points are spread in the RP e.g., 
uniformly distributed noise (Fig~\ref{fig01}C.3),
(ii) periodic and quasi-periodic systems where diagonal lines and 
checkerboard structures represent oscillating systems, e.g., sinusoidal 
signals (Fig~\ref{fig01}D.3),
(iii) drift where paling or darkening recurrence points away from 
the LOI is caused by drifting systems, 
e.g., logistic map (Fig~\ref{fig01}E.3), and
(iv) disrupted where recurrence points are presented white areas or 
bands that indicate abrupt changes in the dynamics, e.g. Brownian motion 
(Fig~\ref{fig01}F.3) \cite{eckmann1987, marwan2015}.
Texture patterns in RPs can be categorised as:
(i) single or isolated recurrence points that represent rare occurring states, 
and do not persist for any time or fluctuate heavily,
(ii) dots forming diagonal lines where the length of the small-scale parallel 
lines in the diagonal are related to the ratio of determinism or predictability 
in the dynamics of the system, and
(iii) dots forming vertical and horizontal lines where the length of the 
lines represent a time length where a state does not change or change very 
slowly and these patterns formation represent discontinuities in the signal, 
and (iv) dots clustering to inscribe rectangular regions which are related 
to laminar states or singularities \cite{marwan2015}.

@@

%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/methods/rps/versions/drawing-v00.pdf}
%    \caption{
%	{\bf Recurrence Plots and its patterns.} 
%	(A) State space of the Lorenz system with controlling parameters 
%	($\rho=28, \sigma=10, \beta=8/3$) where the point, $j$, in trajectory $X()$ which falls into the neighborhood 
%	(black circle) of a given point at $i$ is a recurrent point and is 
%	represented as a black dot in the recurrence plot at location 
%	$(i, j)$ or white otherwise.
%	(B) Recurrence plot using the 
%	three components of the Lorenz system and the RP with no embeddings 
%	and threshold $\epsilon=5$.
%	Time-series with its respective recurrence plot for:
%	(C) uniformly distributed noise,
%	(D) super-positionet harmonic oscillation 
%	($sin( \frac{1}{5}*t) * sin( \frac{5}{100}*t) $),
%	(E) drift logistic map ($x_{i+1} = 4 x_i (1- x_i) $) corrupted 
%	with a linearly increase term ($0.01*i$), and
%	(F) disrupted brownian motion  ($x_{i+1} = x_i + 2*rnorm(1) $).
%	This figure is adapted from \cite{marwan2015}.
%	Code and data to reproduce the figure is available in \cite{srep2021}.
%	}
%    \label{fig:rps}
%\end{figure}
%%%---------------------------------(FIGURE)-------------------------------------

@@

\subsection*{Human-humanoid imitation activities}
For human-humanoid imitation (HHI) activities four neMEMSi sensors were used,
two of which were attached to the right hand of the participant and the 
other two to the left hand of the humanoid robot.
Then, each participant was asked to imitate repetitions of simple horizontal
and vertical arm movements performed by the humanoid robot in the following 
conditions:
(i) ten repetitions of horizontal arm movement at normal (HN) 
	and faster (HF) speed (Figure~\ref{fig:hri} A), and
(ii) ten repetitions of vertical arm movement at normal (VN) 
	and faster (VF) speed (Figure~\ref{fig:hri} C).
The normal and faster speed of arm movements is defined by the duration 
in number of samples of one repetition of NAO's arm movements. 
We select NAO's arm movements duration to distinguish between normal and 
faster arm movements as NAO's movements have less variation 
for such repetitive movements. 
The duration for one repetition of the horizontal 
arm movement at normal speed, HN, is 5 seconds considering 
that each repetition last around 250 samples.
For horizontal arm movement at faster speed, HF, each repetition were performed 
in 2 seconds which correspond to 90 samples of data.
The vertical arm movement at normal speed, VN, were performed  in 6 seconds 
which is around 300 samples of data. For vertical arm movement at 
faster speed, VF, each repetition lasts about 2.4 seconds which correspond 
to 120 samples of data.
To visualise the distinction between normal and faster speed for horizontal 
and vertical arm movements, Fig~\ref{fig:sts} shows smoothed time series 
for axes Z and Y of the gyroscope sensors with four window lengths: 
2-sec (100-samples), 5-sec (250-samples), 10-sec (500-samples) 
and 15-sec (750-samples).
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht] 
\centering
\includegraphics[width=1.0\textwidth]{figures/experiment/sts/pdf/sts}
    \caption{
	{\bf Time series duration of horizontal and vertical arm movements.} 
    		Time series of smoothed data from gyroscope sensor 
		for different speed arm movements performed by NAO: 
		(A) Horizontal Normal arm movement (HN), 
		(B) Horizontal Faster arm movement (HF),
		(C) Vertical Normal arm movement (VN) and 
		(D) Vertical Faster arm movement (VF).
		Additionally, 
		(A) shows window sizes for 2-seconds (100 samples), 
		5-seconds (250 samples), 10-seconds (500 samples) 
		and 15-seconds (750 samples)
		which are also presented in (B), (C) and (D).
		Code and data to reproduce the figure is available in \cite{srep2021}.
        }
	\label{fig:sts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Data collection from inertial measurement units} 
\label{sec:experiment:subsec:imu}
To give insight to the research questions, 
we considered various conditions of time series collected for this work: 
\begin{itemize}
\item Three levels of smoothness for normalised data 
	(e.g., sg0zmuv, sg1zmuv and sg2zmuv) where sg 
	stands for Savitzky-Golay filter with two different filter lengths (29 and 159) 
	and the same polynomial degree of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal}
	and zmuv is zero mean unit variance.
\item four arm movement activities with two velocities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}
%After the data collection, raw time series were windowed, normalised and smoothed. 
Due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig:ts}).
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
    	\caption{ 
	{\bf Time series for horizontal and vertical arm movements.}
		(A/D) raw-normalised (sg0zmuv), 
		(B/E) normalised-smoothed 1 (sg1zmuv) and
		(C/F) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants 
		($p01$, $p02$, and $p03$) 
		for horizontal and vertical arm movements in normal 
		and faster velocity (HN, HF, VN, VF) 
		with the normalised GyroZ or GyroY axis 
		and with one sensor attached to the participant (HS01) 
		and other sensor attached to the robot (RS01).	
	Code and data to reproduce the figure is available in \cite{srep2021}.
        }
    \label{fig:ts}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

@@

%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/caoami/pdf/fig2}
	\caption{
	{\bf Box plots of minimum embedding parameters.} 
	Box plots of (A) minimum embedding dimensions 
	and (B) first minimum AMI values for 
	Horizontal Normal (HN), Horizontal Faster (HF),
	Vertical Normal (VN) and Vertical Faster (VF)
	with sensors attached to participants (HS01) and
	sensor attached to robot (RS01).
	Minimum embedding parameters are for twenty participants 
	($p01$ to $p20$) with three smoothed signals 
	(sg0: sg0zmuvGyroZ, sg1: sg1zmuvGyroZ and sg2: sg2zmuvGyroZ)
	and window length of 10-sec (500 samples).
	Code and data to reproduce the figure is available in \cite{srep2021}.
        }
    \label{fig:cao_ami}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsubsection*{Uniform Time-Delay Embedding}
Using the overall embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$), 
the first three axis of the rotated Principal Components Analysis (PCA) 
are shown for the reconstructed state spaces of horizontal (figures~\ref{fig:rsss}(A)) 
and vertical (figures~\ref{fig:rsss}(B)) arm movements. 
While visual inspection of these figures 
suggests differences in the trajectories of the reconstructed state spaces, 
we require an objective quantification to determine the extent of the differences.  
One approach could be use Euclidean distances from the origin for points in these 
trajectories, but this proved inconclusive and was not able to capture 
the dynamics of the trajectories. Consequently, we applied 
Recurrence Plots and Recurrence Quantification Analysis.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/versions/drawing-v00.pdf}
\caption{
	{\bf RSSs for horizontal and vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$
	for (A) horizontal and (B) vertical arm movements.
	Code and data to reproduce the figure is available in \cite{srep2021}.	
        }
    \label{fig:rsss}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrences Plots}
Using the average embedding parameters 
($\overline{m}_0=6$, $\overline{\tau}_0=8$) 
and an recurrence threshold of $\epsilon=1$.
As our interest is for dynamical transitions, 
there is little importance on the selection of $\epsilon$ which in this case
is 1. 
Recurrence Plots (RP) were computed for horizontal (figures~\ref{fig:rpsr}(A)) and 
vertical (figures~\ref{fig:rpsr}(B)) arm movements.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/versions/drawing-v00.pdf}
\caption{
	{\bf RPs for horizontal (A) and vertical (B) arm movements.}	
	Recurrence plots were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$.
	Code and data to reproduce the figure is available in \cite{srep2021}.
        }
    \label{fig:rpsr}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrence Quantification Analysis} \label{ch6:rqas}
Four Recurrence Quantification Analysis metrics (
percentage recurrence, REC, representing the percentage of black dots in RP; 
percentage determinism, DET, representing the predictability of the RP; 
ratio of DET / REC, RATIO; Shannon entropy, ENTR) 
were computed using the same parameters as for RP.
%\subsubsection*{REC values}
Figure~\ref{fig:RQABP}(A) presents box plots of REC values, 
for HS01, are more spread for Slow (i.e., 5 seconds per movement) movements 
in Horizontal (HS) or Vertical (VS) than for Fast (i.e., 2 seconds per movement).  
This suggests greater variation between participants for the Slow movements.  
For RS01, there is little variation between Slow and Fast movement
(interquartile range of 0.01). 
In terms of smoothness, there seems little effect of HS01 but RS01 values 
do show affects of smoothness (see the incremental changes of mean values (rhombus)).
%\subsubsection*{DET values}
Figure \ref{fig:RQABP}(B) presents DET values and shows 
little difference for type of movement or performer.  
However, DET values are affected by changes in smoothness of the signal, 
particularly for Fast movement.
%\subsubsection*{RATIO values}
Figure \ref{fig:RQABP}(C) presents the ratio of DET / REC. 
These values, for the human performer, vary less for HN than for HF movement.  
Additionally, smoothness leads to a decrease in mean values for Fast movements.
%\subsubsection*{ENTR values}
Figure \ref{fig:RQABP}(D) shows ENTR values are higher for the human 
performer than the robot, and vary with the smoothness of the time-series.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/rqa/output/rqa-bp}
    \caption
	[Box plots for RQA values]{
	{\bf Box plots for RQA metrics.}
	RQA metrics for (A) REC, (B) DET, (C) RATIO, and (D) ENTR of 
	20 participants performing HN, HF, VN and VF movements
	with sensors HS01, RS01 and three smoothed-normalised  
	time series (sg0, sg1 and sg2).
	RQA values were computed with 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$. 
	Code and data to reproduce the figure is available in \cite{srep2021}.
        }
    \label{fig:RQABP}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
